\chapter{Euclidean algorithm}
\label{ch:euclid}

The goal of this chapter is to study the \term{Euclidean algorithm},
which is an algorithm for quickly finding the greatest common divisor
of two numbers.

First, we're going to go over some basics, starting with some
inductive proofs about properties of the natural numbers.

\input{chapters/1/1.tex}
\input{chapters/1/2.tex}
\input{chapters/1/3.tex}
% \input{chapters/1/4.tex}

\section*{Exercises}

\subsection*{Induction}

\begin{exercise}
  \label{exc:strong-weak}
  Prove that strong induction \parenref{def:strong-induction}
  implies weak induction \parenref{def:weak-induction}.
\end{exercise}

\begin{exercise}
  \label{exc:well-ordering}
  Prove that the well-ordering
  principle \parenref{def:well-ordering} is equivalent to
  mathematical induction \parenref{def:strong-induction}.
\end{exercise}

\exercise{Prove \cref{thm:n-positive-ints} without induction.}

\begin{exercise}
  \label{exc:fermat}
  \label{fermat}
  Using induction, prove that, for all natural numbers $n > 2$,
  there are no integer solutions to

  \begin{zz}
    x^n + y^n = z^n
  \end{zz}
\end{exercise}

\subsubsection*{Aaron's Induction Exercises}

\begin{exercise}
  \label{aaron-induction-1}
  Using induction, prove each of the following

  \begin{enumerate}[label=(\alph*)]
  \item For $n \ge 1$, the sum of the first $n$ positive integers is
    equal to $\frac{n(n + 1)}{2}$. That is,

    \begin{zz}
      1 + 2 + 3 + \dots + n = \frac{n(n + 1)}{2}
    \end{zz}
  \item For $n \ge 1$, sum of the first $n$ perfect squares is equal
    to $\frac{n(n + 1)(2n + 1)}{6}$. That is,

    \begin{zz}
      1^2 + 2^2 + 3^2 + \dots + n^2 = \frac{n (n + 1)(2n + 1)}{6}
    \end{zz}
  \item Using induction, prove that for $n \ge 1$, the sum of the
    first $n$ positive cubes is equal to the square of the sum of
    the first $n$ positive integers. That is,

    \begin{zz}
      1^3 + 2^3 + 3^3 + \dots + n^3 = \parens{1 + 2 + 3 + \dots + n}^2
    \end{zz}

    Hint: use part (a).
  \item Fix a real number $x \ge -1$. For any $n \ge 1$,

    \begin{zz}
      (1 + x)^n \ge 1 + nx
    \end{zz}

  \item For $n \ge 1$,

    \begin{zz}
      1(2) + 2(3) + 3(4) + \dots + n(n + 1) = \frac{n(n+1)(n+2)}{3}
    \end{zz}
  \item For $n \ge -1$, the integer $2n^3 + 7n$ is divisible by $3$.
  \item For $n \ge 1$, the integer $n^2 + (n + 2)^2 + (n + 4)^2$ is
    divisible by $12$.
  \end{enumerate}
\end{exercise}

\begin{exercise}
  Compute the first several values of the following sequences, and
  look for a pattern. Guess the formula for the $n$th term. Prove your
  guess is correct using induction.

  \begin{enumerate}[label=(\alph*)]
  \item $a_n = \sum_{k = 1}^n 2^{-k}$
  \item $b_n = \sum_{k = 1}^n \frac{1}{k(k + 1)}$
  \end{enumerate}
\end{exercise}

\begin{exercise}
  The Fibonacci sequence $F_n$ is defined recursively by

  \begin{rcl}
    F_0 & = & 0 \\
    F_1 & = & 1 \\
    F_n & = & F_{n - 1} + F_{n - 2} \\
  \end{rcl}

  \begin{enumerate}[label=(\alph*)]
  \item Calculate $F_n$ for $2 \le n \le 15$.
  \end{enumerate}

  Use induction to prove the following statements

  \begin{enumerate}[label=(\alph*),start=2]
  \item For $n \ge 0$,

    \begin{zz}
      F_1 + F_2 + F_3 + \dots + F_n = F_{n + 2} - 1
    \end{zz}

  \item For $n \ge 1$,

    \begin{zz}
      F_1 + F_3 + F_5 + \dots + F_{2n - 1} = F_{2n}
    \end{zz}

  \item for $n \ge 1$,

    \begin{zz}
      F_2 + F_4 + F_6 + \dots + F_{2n} = F_{2n + 1} - 1
    \end{zz}

  \item Use induction to prove that for $n \ge 0$,

    \begin{zz}
      \parens{F_{n + 1}}^2 - F_nF_{n + 2} = (-1)^n
    \end{zz}
    
  \end{enumerate}
\end{exercise}

\subsection*{Division}

\subsubsection*{Aaron's Divisibility exercises}

\begin{exercise}
  Prove the following statements about the integers $a, b, c, d$

  \begin{enumerate}[label=(\alph*)]
  \item If $a \div 1$, then $a = \pm 1$.
  \item If $a \div b$, and $b \div a$, then $a = \pm b$
  \item If $a \div b$, and $c \div d$, then $ab \div cd$
  \item If $a \div b$, and $a \div c$, then $a^2 \div bc$
  \item If $a \div b$, and $c \ne 0$, then $ac \div bc$
  \item If $ac \div bc$, and $c \ne 0$, then $a \div b$
  \end{enumerate}
\end{exercise}

\begin{exercise}
  Disprove the following statements about positive integers $a$, $b$,
  and $c$ by finding explicit counterexamples:

  \begin{enumerate}[label=(\alph*)]
  \item If $a \div (b + c)$, then $a \div b$ and $a \div c$.
  \item If $a \div bc$, then $a \div b$ or $a \div c$.
  \end{enumerate}
\end{exercise}

\begin{exercise}
  Prove the following statements

  \begin{enumerate}[label=(\alph*)]
  \item For any integer $n$, exactly one of $n$, $n + 2$, or $n + 4$
    is divisible by $3$
  \item For any integer $n$, the integer $2n^3 + 7n$ is divisible by
    $3$.
  \item The square of any even integer is divisible by $4$.
  \item The square of any odd integer is of the form $8n + 1$.
  \item For any odd integer $n$, the integer
    $n^2 + (n + 2)^2 + (n + 4)^2 + 1$ is divisible by $12$.
  \item For any integer $n$, the integer $(n + 1)^3 - n^3$ is odd.
  \end{enumerate}
\end{exercise}

\begin{exercise}
  \label{exc:gcds}
  Use the Euclidean algorithm to find the greatest common divisors
  below.

  \begin{enumerate}[label=(\alph*)]
  \item $\gcdp{339, 82}$
  \item $\gcdp{8415, 1078}$
  \item $\gcdp{1989, 1881}$
  \end{enumerate}
\end{exercise}

\begin{exercise}
  \label{exc:cfs}
  Use the continued fraction algorithm to write the following rational
  numbers as continued fractions. Then, compute the second-to-last
  convergent of the continued fraction.

  \begin{enumerate}[label=(\alph*)]
  \item $\frac{339}{82}$
  \item $\frac{8415}{1078}$
  \item $\frac{1989}{1881}$
  \end{enumerate}
\end{exercise}

\begin{exercise}
  \label{exc:intsolutions}
  Use the Euclidean algorithm output from \cref{exc:gcds} to find
  integer solutions to the following equations. Compare these
  solutions to the second-to-last convergents calculated in
  \cref{exc:cfs}.

  \begin{enumerate}[label=(\alph*)]
  \item $339x + 82y = \gcdp{339, 82}$
  \item $8415x + 1078y = \gcdp{8415, 1078}$
  \item $1989x + 1881y = \gcdp{1989, 1881}$
  \end{enumerate}
\end{exercise}

\begin{exercise}
  Find \xtb{all} integer solutions to the equations in
  \cref{exc:intsolutions}.
\end{exercise}